---
title: "Reproducible Research: Peer Assessment 1"
author: "Rubina Sattar"
date: "Sunday, November 16, 2014"
output: html_document
---

Loading and preprocessing the data

```{r}
setwd("C:/Users/Rubina/Desktop/R/Class/RR/RepData_PeerAssessment1")
unzip(zipfile = "activity.zip")
data <- read.csv("activity.csv", header= TRUE)
```
### What is mean total number of steps taken per day?
##### 1. Histogram of the total number of steps taken each day

```{r}
library(ggplot2)
total.steps <- tapply(data$steps, data$date, FUN = sum, na.rm = TRUE)
qplot(total.steps, binwidth = 1000, xlab = "total number of steps taken each day")
```

##### 2. Calculate and report the mean and median total number of steps taken per day
```{r}
mean(total.steps, na.rm = TRUE)
median(total.steps, na.rm = TRUE)
```

### What is the average daily activity pattern?
##### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
library(ggplot2)
averages <- aggregate(x = list(steps = data$steps), by = list(interval = data$interval), 
    FUN = mean, na.rm = TRUE)
ggplot(data = averages, aes(x = interval, y = steps)) + geom_line() + xlab("5-minute interval") + 
    ylab("average number of steps taken")
```

##### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
averages[which.max(averages$steps), ]
```

### Imputing missing values

##### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
missing <- is.na(data$steps)
# How many missing values ?
table(missing)
```

##### 2. Devise a strategy for filling in all of the missing values in the datase. Replacing each missing value with the mean value of its 5-minute interval:
```{r}
fill.value <- function(steps, interval) {
    filled <- NA
    if (!is.na(steps)) 
        filled <- c(steps) else filled <- (averages[averages$interval == interval, "steps"])
    return(filled)
}
filled.data <- data
filled.data$steps <- mapply(fill.value, filled.data$steps, filled.data$interval)
```

##### 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
total.steps <- tapply(filled.data$steps, filled.data$date, FUN = sum)
qplot(total.steps, binwidth = 1000, xlab = "total number of steps taken each day")
```

##### 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
mean(total.steps)
median(total.steps)
```

##### Mean and median values are higher after imputing missing data. The reason is that in the original data, there are some days with steps values NA for any interval. The total number of steps taken in such days are set to 0s by default. However, after replacing missing steps values with the mean steps of associated interval value, these 0 values are removed from the histogram of total number of steps taken each day.  As a result, mean and median are higher and exactly the same with the missing data filled in.

### Differences in activity patterns between weekdays and weekends
##### Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day (Using the weekdays() function and the filled-in data set )
```{r}
weekday.or.weekend <- function(date) {
    day <- weekdays(date)
    if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")) 
        return("weekday") else if (day %in% c("Saturday", "Sunday")) 
        return("weekend") else stop("invalid date")
}
filled.data$date <- as.Date(filled.data$date)
filled.data$day <- sapply(filled.data$date, FUN = weekday.or.weekend)
```

##### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r}
averages <- aggregate(steps ~ interval + day, data = filled.data, mean)
ggplot(averages, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) + 
    xlab("5-minute interval") + ylab("Number of steps")
```

